Estimating the dynamics of\\SPARQL query results over Wikidata\thanks{This work was supported by CONICYT PFCHA, Doctorado Becas Chile/2017 -- 72180000, the Millennium Institute for Foundational Research on Data (IMFD) and FONDECYT Grant No. 1181896.}

### 2019-01-26 kgswc19.tex
Finished pass up to end of Section 5

\colorlet{punct}{red!60!black}
\definecolor{background}{HTML}{EEEEEE}
\definecolor{delim}{RGB}{120,20,40}
\definecolor{keyw}{RGB}{0,0,192}
\colorlet{numb}{magenta!60!black}

\lstdefinelanguage{sparql}{
	sensitive=false,
	extendedchars=true,
	literate={á}{{\'a}}1 {é}{{\'e}}1 {í}{{\'{\i}}}1 {ó}{{\'o}}1 {ú}{{\'u}}1
	{Á}{{\'A}}1 {É}{{\'E}}1 {Í}{{\'I}}1 {Ó}{{\'O}}1 {Ú}{{\'U}}1
	{ü}{{\"u}}1 {Ü}{{\"U}}1 {ñ}{{\~n}}1 {Ñ}{{\~N}}1 {¿}{{?``}}1 {¡}{{!``}}1
	{<}{{{\color{delim}<}}}{1}
	{>}{{{\color{delim}>}}}{1}
	{?}{{{\color{delim}?}}}{1}
	{*}{{{\color{delim}*}}}{1}
	{+}{{{\color{delim}+}}}{1}
	{/}{{{\color{delim}/}}}{1}
	{,}{{{\color{punct}{,}}}}{1}
	{;}{{{\color{punct}{;}}}}{1}
	{.}{{{\color{punct}{.}}}}{1}
	{:}{{{\color{punct}{:}}}}{1}
	{\{}{{{\color{delim}{\{}}}}{1} {\}}{{{\color{delim}{\}}}}}{1},
	morekeywords={ask,select,from,where,order,by,distinct,limit,offset,optional,union,filter,prefix,bound,desc,regex,str,group,not,exists,minus,service,certain,maybe}
}

\lstdefinestyle{sparqld}{
	basicstyle=\scriptsize\ttfamily,
	identifierstyle=\color{black},
	keywordstyle=\color{keyw}\bfseries,
	ndkeywordstyle=\color{greenCode}\bfseries,
	stringstyle=\color{ocherCode}\ttfamily,
	commentstyle=\color{darkgray}\ttfamily,
	language={sparql},
	tabsize=2,
	showtabs=false,
	showspaces=false,
	showstringspaces=false,
	extendedchars=true,
	escapechar=`,
	frame={single},
	breaklines=true,
	basewidth=0.5em,
	moredelim=[is][\color{magenta}]{~}{~},
	moredelim=**[is][\color{gray}]{£}{£},
	moredelim=**[is][\color{blue!50!black}]{$}{$},
	moredelim=**[is][\color{green!50!black}]{¬}{¬},
	xleftmargin=2ex,
	xrightmargin=1ex,
	aboveskip=1.5ex,
	belowskip=1.5ex
}

@@

\paragraph{Architecture:} We propose a method to predict answers for the OSC and TTL problems. Our method primarily considers static features of the query: these include statistics such as the number of triple patterns, as well as the presence of query operators, such as \texttt{UNION}, \texttt{OPTIONAL}, specific path expressions, etc.; our model can also consider statistics about predicates in the dynamic RDF graph, as well as in which past versions of the dynamic graph the results for the query changed. Figure~\ref{fig:schema} shows the components of our approach, which receives as input a SPARQL query (the architecture itself already encodes information about the dynamic graph and the value of $k$ for OSC, which we set in experiments to 1, predicting changes in the next version). The Feature Extractor decomposes the query into a set of numeric features. The Dynamics Model accepts some features of the query, as well as information from the dynamic graph, in order to compose further features. The final features are passed to a classifier in the case of the OSC problem, and to a regression model in the case of the TTL problem, to generate the final output prediction. We now describe the Feature Extractor and Dynamics Model in more detail.

@@

The final feature -- a raw set of all predicates used -- is rather a meta-feature that serves as input for the next part of the proposed model.

@@

\begin{figure}[!t]
{\centering
\hfill
	\begin{minipage}{0.35\textwidth}
	\begin{lstlisting}[style=sparqld]
:LMessi :instance_of :Human ;
 :children 1 , 2 ;
 :played :2014WC .
	\end{lstlisting}
	\end{minipage}	
\hfill
\begin{minipage}{0.35\textwidth}
	\begin{lstlisting}[style=sparqld]
:LMessi :instance_of :Human ;
 :children 2 , 3 .
	\end{lstlisting}
\end{minipage}	
\hfill
}

\caption{Changes for $\mathcal{G}$ in Figure~\ref{fig:dgc}, showing $G_1 \oplus G_2$ (left) and $G_2 \oplus G_3$ (right) \label{fig:dgc}}
\end{figure}

@@

@@

We conducted an empirical study to evaluate the effectiveness of our proposal to predict change in SPARQL query results. For this, we need an RDF dataset, with a lot of information, preferably, cross-domain, and that is constantly updated since we assume that the information of the dataset is correct, complete, but dynamic. We also need a set of queries appropriate to the chosen dataset, preferably a heterogeneous set defined by real users and covering a large part of the dataset.

\subsection{RDF Dataset}

We use 23 Wikidata snapshots from 18/04/2017 to 27/09/2017, which are captured weekly in truthy version, which represents statements that have the best non-deprecated rank for a given property. 
The first version has 1,102,242,331 triples, 57,499,191 unique IRIs, 3,276 unique predicates, 933,588,860 literals, and 7,449 blank nodes, while the final version has 1,924,967,162 triples (+74\%), 80,637,164 unique IRIs (+40\%), 3,660 unique predicates (+11\%), 1,676,930,476 literals (+79\%) and 8,909 blank nodes (+19\%). Figure~\ref{graph:datasize} shows the growth maintained over the period of some of these indicators. The gap in week 12 is due to the fact that we were not able to obtain the data for that week. Even though the information is relatively incremental, there are deletions. Figure~\ref{graph:delta} shows the ratio of triples eliminated and added each week.

\begin{figure}[h]
	\begin{minipage}[b]{0.40\linewidth}
		\centering
		\begin{tikzpicture}[thick, scale=0.5]
		\begin{axis}[
		%title={Temperature dependence of CuSO$_4\cdot$5H$_2$O solubility},
		xlabel={Version},
		ymin=0,
		xtick={0,5,10,15,20,25},
		legend pos=north west,
		ymajorgrids=true,
		grid style=dashed,
		]

@@

		\end{axis}
		\end{tikzpicture}
		\caption{Evolution of the number of triples on Wikidata.}
		\label{graph:datasize}
	\end{minipage}
	\hspace{0.5cm}
	\begin{minipage}[b]{0.40\linewidth}
		\centering
		\begin{tikzpicture}[thick, scale=0.5]
		\begin{axis}[
		x tick label style={
			/pgf/number format/1000 sep=},
		%ylabel=\#Triples,
		enlargelimits=0.05,
		legend style={at={(0.5,-0.1)},
			anchor=north,legend columns=-1},
		ybar interval=0.7,
		]
		\addplot 
		coordinates {(1,40137188)(2,4563382)(3,29302968)(4,8906402)(5,11455047)(6,11435682)(7,27229516)(8,26968577)(9,38952400)(10,20283316)(11,0)(12,0)(13,70073220)(14,17685830)(15,45405782)(16,41199256)(17,42633484)(18,56683899)(19,25309217)(20,25030143)(21,22548371)(22,23330356)(23,37009794)};
		\addplot 
		coordinates {(1,3403743)(2,2235268)(3,1116111)(4,3096270)(5,3371448)(6,5156519)(7,5873153)(8,1336452)(9,6744349)(10,999601)(11,0)(12,0)(13,1123311)(14,4002828)(15,7250907)(16,5356971)(17,6495717)(18,5627376)(19,4812243)(20,790590)(21,681118)(22,1194899)(23,998105)};
		\legend{Added, Removed}
		\end{axis}
		\end{tikzpicture}
		\caption{Triples added and removed in Wikidata.}
		\label{graph:delta}
	\end{minipage}
\end{figure}

\subsection{Queries Dataset}

As opposed to RDF data, which can be easily obtained in the form of dumps, at the time of conducting this research, we did not find any structured dataset with queries following the Wikidata data model and Wikidata query logs where inaccessible. We considered then 388 queries of examples from Wikidata page\footnote{https://www.wikidata.org/wiki/Wikidata:SPARQL\_query\_service/queries/examples}.

Then, the queries had to be curated manually, because they had syntax errors; others asked information external to Wikidata or asked for qualifiers that are not present in the truthy version. We also eliminated the queries that returned bindings to blank nodes to facilitate the comparison between the results. Therefore, we need deterministic queries. In our case, we eliminate the queries with the keyword SAMPLE and add ORDER BY to all the queries with all the projected variables. The latter change also facilitates the detection of changes in the results. Also, we removed the use of custom Label service of Wikidata\footnote{https://www.mediawiki.org/wiki/Wikidata\_Query\_Service/User\_Manual\#Label\_service}.

Figures~\ref{fig:languageB} and \ref{fig:languageA} shows an example of several transformations of a query.

\begin{lstlisting}[captionpos=b, caption=Query transformation. Before., label=fig:languageB,
basicstyle=\ttfamily,frame=single]
#Cats
SELECT ?item ?itemLabel
WHERE {
  ?item wdt:P31 wd:Q146 .
  SERVICE wikibase:label
    { bd:serviceParam wikibase:language "en,es" }
}
\end{lstlisting}

\begin{lstlisting}[captionpos=b, caption=Query transformation. After., label=fig:languageA,
basicstyle=\ttfamily,frame=single]
#Cats
SELECT ?item ?itemLabel
WHERE {
  ?item wdt:P31 wd:Q146 .
  OPTIONAL {?item rdfs:label ?en.FILTER(LANG(?en)="en")}
  OPTIONAL {?item rdfs:label ?es.FILTER(LANG(?es)="es")}
  BIND(str(COALESCE(?en, ?es, strafter(str(?item),
    "http://www.wikidata.org/entity/"))) AS ?itemLabel)
}
ORDER BY ?item ?itemLabel
\end{lstlisting}

In the end, we have 221 curated queries. Then, we analyze the structure of the triple patterns of the queries. This is important, because our notion of freshness focuses on predicates. This restricts our approach to queries with triple patterns with a constant in the predicate position. Table~\ref{tab:Pattern} shows that in total 90.35\% of the triple patterns have a constant in the predicate position, which validate our approach of using a model based on the dynamics of predicates.

\begin{table}[h]
	\begin{minipage}[b]{0.45\linewidth}
		\caption{Triple patterns (C=Constant, V=Variable).}
		\begin{tabular}{lrr}    \hline
			Pattern & Absolute & Relative \\    \hline
			V C V   & 751      & 71.73\%    \\
			V C C   & 189      & 18.05\%    \\
			V V V   & 9        & 0.86\%     \\
			C C V   & 6        & 0.57\%     \\
			V V C   & 5        & 0.48\%     \\
			C V V   & 3        & 0.29\%     \\
			C V C   & 1        & 0.10\%      \\
			C C C   & 0        & 0.00\%        \\    \hline
			V P C   & 69       & 6.59\%     \\
			V P V   & 13       & 1.24\%     \\
			C P V   & 1        & 0.10\%      \\
			C P C   & 0        & 0.00\%        \\    \hline
		\end{tabular}
		\label{tab:Pattern}
	\end{minipage}
	\hspace{1cm}
	\begin{minipage}[b]{0.45\linewidth}
		\caption{Structure of navigational Property Paths.}
		\label{tab:Path}
		\begin{tabular}{lrr}    \hline
			Rules                & Absolute & Relative \\    \hline
			$P \rightarrow P* $   & 62       & 74.70\%     \\
			$P \rightarrow P / P$  & 56       & 67.47\%    \\
			$P \rightarrow P+ $   & 8        & 9.64\%     \\
			$P \rightarrow P \mid P$  & 6        & 7.23\%     \\
			$P \rightarrow P? $   & 6        & 7.23\%     \\
			$P \rightarrow !(P)$ & 0        & 0.00\%        \\
			$P \rightarrow C $    & 83       & 100.00\%      \\    \hline
		\end{tabular}
	\end{minipage}
\end{table}

Table~\ref{tab:Pattern} also shows that 7.93\% of triple patterns contain a Property Path in the predicate position. Then we set out to analyze these triples more deeply. Table~\ref{tab:Path} shows the frequency of occurrence of these patterns in our queries. It shows that $ P \rightarrow P * $ (74.70\%) and $ P \rightarrow P + $ (9.64\%) are often applied, which corresponds to our intuition to have a feature that indicates if there are triple patterns offering arbitrary length paths (*PATH). We should also notice that there is no negation at the level of the path. Our feature extractor is able to extract all constants of predicates within paths.

Finally, we evaluate the dynamics of the results of the queries considering Equation~\ref{eq:delta} between query results of the same query over consecutive versions of Wikidata.

Figure~\ref{graph:changeV} shows the sum of the $ \delta $ function for all the queries considering the consecutive results of each query for 23 weeks. We identify that approximately half of the results of the queries change every week.

Figure~\ref{graph:changeH} shows for each query how many changes registered over 23 weeks. We identify queries that change every week, others that never changed and a fairly uniform distribution of changes in the period.

\begin{figure}[h]
	\begin{minipage}[b]{0.40\linewidth}
		\centering
		\begin{tikzpicture}[thick, scale=0.5]
		\begin{axis}[
		%title={Temperature dependence of CuSO$_4\cdot$5H$_2$O solubility},
		xlabel={Version},
		ylabel={Changes},
		ymin=0, ymax=221,
		xtick={0,5,10,15,20,25},
		legend pos=north west,
		ymajorgrids=true,
		grid style=dashed,
		]

@@

		\end{axis}
		\end{tikzpicture}
		\caption{Query results changes by time.}
		\label{graph:changeV}
	\end{minipage}
	\hspace{0.7cm}
	\begin{minipage}[b]{0.40\linewidth}
		\centering
		\begin{tikzpicture}[thick, scale=0.5]
		\begin{axis}[
		%title={Temperature dependence of CuSO$_4\cdot$5H$_2$O solubility},
		xlabel={Queries},
		ymin=0,
		legend pos=north west,
		ymajorgrids=true,
		grid style=dashed,
		]

@@

		\end{axis}
		\end{tikzpicture}
		\caption{Query results changes by queries.}
		\label{graph:changeH}
	\end{minipage}
\end{figure}

>>>>

\colorlet{punct}{red!60!black}
\definecolor{background}{HTML}{EEEEEE}
\definecolor{delim}{RGB}{120,20,40}
\definecolor{keyw}{RGB}{0,0,192}
\definecolor{stri}{RGB}{100,0,130}
\colorlet{numb}{magenta!60!black}

\lstdefinelanguage{sparql}{
	sensitive=false,
	extendedchars=true,
	literate={á}{{\'a}}1 {é}{{\'e}}1 {í}{{\'{\i}}}1 {ó}{{\'o}}1 {ú}{{\'u}}1
	{Á}{{\'A}}1 {É}{{\'E}}1 {Í}{{\'I}}1 {Ó}{{\'O}}1 {Ú}{{\'U}}1
	{ü}{{\"u}}1 {Ü}{{\"U}}1 {ñ}{{\~n}}1 {Ñ}{{\~N}}1 {¿}{{?``}}1 {¡}{{!``}}1
	{<}{{{\color{delim}<}}}{1}
	{>}{{{\color{delim}>}}}{1}
	{?}{{{\color{delim}?}}}{1}
	{*}{{{\color{delim}*}}}{1}
	{+}{{{\color{delim}+}}}{1}
	{/}{{{\color{delim}/}}}{1}
	{,}{{{\color{punct}{,}}}}{1}
	{;}{{{\color{punct}{;}}}}{1}
	{.}{{{\color{punct}{.}}}}{1}
	{:}{{{\color{punct}{:}}}}{1}
	{\{}{{{\color{delim}{\{}}}}{1} {\}}{{{\color{delim}{\}}}}}{1}
	{(}{{{\color{punct}(}}}{1} {)}{{{\color{punct})}}}{1},
	morekeywords={ask,select,from,where,order,by,distinct,limit,offset,optional,union,filter,prefix,bound,desc,lang,regex,str,group,not,exists,minus,service,certain,maybe,coalesce,bind,as,strafter}
}

\lstdefinestyle{sparqld}{
	basicstyle=\scriptsize\ttfamily,
	identifierstyle=\color{black},
	keywordstyle=\color{keyw}\bfseries,
	ndkeywordstyle=\color{greenCode}\bfseries,
	stringstyle=\color{stri}\ttfamily,
	commentstyle=\color{gray}\ttfamily,
	language={sparql},
	tabsize=2,
	showtabs=false,
	showspaces=false,
	showstringspaces=false,
	extendedchars=true,
	escapechar=`,
	frame={single},
	breaklines=true,
	basewidth=0.5em,
	comment=[l]{\#},
	morestring=[b]",
	moredelim=[is][\color{magenta}]{~}{~},
	moredelim=**[is][\color{gray}]{£}{£},
	moredelim=**[is][\color{blue!50!black}]{$}{$},
	moredelim=**[is][\color{green!50!black}]{¬}{¬},
	xleftmargin=2ex,
	xrightmargin=1ex,
	aboveskip=1.5ex,
	belowskip=1.5ex
}

@@

\paragraph{Architecture:} We propose a method to predict answers for the OSC and TTL problems. Our method primarily considers static features of the query: these include statistics such as the number of triple patterns, as well as the presence of query operators, such as \texttt{UNION}, \texttt{FILTER}, specific path expressions, etc.; our model can also consider statistics about predicates in the dynamic RDF graph, as well as in which past versions of the dynamic graph the results for the query changed. Figure~\ref{fig:schema} shows the components of our approach, which receives as input a SPARQL query (the architecture itself already encodes information about the dynamic graph and the value of $k$ for OSC, which we set in experiments to 1, predicting changes in the next version). The Feature Extractor decomposes the query into a set of numeric features, described presently. The Dynamics Model accepts some features of the query, as well as information from the dynamic graph, in order to compose further features. The final features are passed to a classifier in the case of the OSC problem, and to a regression model in the case of the TTL problem, to generate the final output prediction. We now describe the Feature Extractor and Dynamics Model in more detail.

@@

The final feature is the set of all predicates used in a triple pattern, including (where applicable) the IRIs used in a path expression; this is rather a meta-feature that serves as input for the next part of the proposed model.

@@

\begin{figure}[!t]
{\centering
\hfill
	\begin{minipage}{0.35\textwidth}
	\begin{lstlisting}[style=sparqld]
:LMessi :instance_of :Human ;
 :children 1 , 2 ;
 :played :2014WC .
	\end{lstlisting}
	\end{minipage}	
\hfill
\begin{minipage}{0.35\textwidth}
	\begin{lstlisting}[style=sparqld]
:LMessi :instance_of :Human ;
 :children 2 , 3 .
	\end{lstlisting}
\end{minipage}	
\hfill
}
\caption{Changes for $\mathcal{G}$ in Figure~\ref{fig:dgc}, showing $G_1 \oplus G_2$ (left) and $G_2 \oplus G_3$ (right) \label{fig:dgc}}
\end{figure}

@@

Given a query $Q$ with a set of predicates $\{ p_i, ..., p_n \}$, we may then consider a variety of aggregate functions over $\Delta(\mathcal{G},p_1), ..., \Delta(\mathcal{G},p_n)$, such as $\mathrm{max}$, $\mathrm{mean}$, etc., to compute a final numeric feature for the query, representing a summary of the level of dynamicity of the predicates it contains.

@@

In order to evaluate the effectiveness of our proposal for predicting changes in SPARQL query results, we require a dynamic RDF graph, with access to various historical versions; preferably this graph contains real-world, large-scale, diverse data, and with sufficient changes between versions to provide both positive and negative examples of queries whose results change. Furthermore, we require a set of SPARQL queries that can be answered against this dynamic graph; preferably these queries again should be diverse, representative of real-world user queries, of a variety of shapes and sizes, using a variety of query operators, and with a mix of both dynamic and static results over the dynamic graph. In particular, we choose Wikidata~\cite{VrandecicK14} for our experiments, which, we shall argue, meets the aforementioned requirements. We first give details of the dynamic data we collect from Wikidata; thereafter, we discuss the queries we use in our experiments.

\paragraph{RDF Data:} We use 23 Wikidata snapshots from 18/04/2017 to 27/09/2017, which are captured weekly in the \emph{truthy version} that contains triples (without qualifiers) that have the best non-deprecated rank for a given property. The first version has 1,102,242,331 triples and 3,276 unique predicates, while the final version has 1,924,967,162 triples (+74\%) and 3,660 unique predicates (+11\%). Figure~\ref{fig:wikidata}, on the left, shows the growth in triples as the weeks progress, while on the right, shows the numbers of triples added and removed week-to-week. We see that although many more triples are added than removed, there are some triples removed each week. We further note some peaks in triples added in some weeks, which may be due to bulk imports of data. The gap in week 12 is due to the fact that we were not able to obtain the data for that week (we thus exclude the triples added/removed from weeks 11 to 12, and 12 to 13).

\ah{Regarding: ``The first version has 1,102,242,331 triples, 57,499,191 unique IRIs, 3,276 unique predicates, 933,588,860 literals, and 7,449 blank nodes, while the final version has 1,924,967,162 triples (+74\%), 80,637,164 unique IRIs (+40\%), 3,660 unique predicates (+11\%), 1,676,930,476 literals (+79\%) and 8,909 blank nodes (+19\%).'' I assume that the literals and blank nodes are not unique? Also the IRIs are unique per position? (In any case, such data is not 100\% needed; what's more important are the unique triples and unique predicates.)}

\ah{Would it be possible to get a table of the top-10 most and least dynamic predicates by number of triples added/removed over the weeks, and also the number of triples with each such predicate in the first and last week?}

\begin{figure}[t]
	%\begin{minipage}[b]{0.47\linewidth}
		\centering
		\begin{tikzpicture}[thick,font=\scriptsize]
		\begin{axis}[
		%title={Temperature dependence of CuSO$_4\cdot$5H$_2$O solubility},
		xlabel={Version},
		width=5cm, 
		height=5cm,
		ymin=0,
		xtick={1,2,...,23},
		xticklabels={},
		extra x ticks={1,23},
		legend pos=north west,
		ymajorgrids=true,
		grid style=dashed,
		]

@@

		\end{axis}
		\end{tikzpicture}
%		\caption{Evolution of the number of triples on Wikidata.}
%		\label{graph:datasize}
%	\end{minipage}
	\hspace{0.1cm}
%	\begin{minipage}[b]{0.47\linewidth}
%		\centering
		\begin{tikzpicture}[thick,font=\scriptsize]
		\begin{axis}[
		xlabel={Version},
		x tick label style={
			/pgf/number format/1000 sep=},
		%ylabel=\#Triples,
		width=9cm, 
		height=5cm,
		enlarge x limits=0.04,
		enlarge y limits=0,
		ymax=80000000,
		legend style={at={(0.25,0.95)},
			anchor=north,legend columns=-1},
		ybar=0pt,
		xtick=data,
		xtick align=inside,
		bar width=3pt,
		ymajorgrids,
		grid style=dashed,
		]
		\addplot 
		coordinates {(1,40137188)(2,4563382)(3,29302968)(4,8906402)(5,11455047)(6,11435682)(7,27229516)(8,26968577)(9,38952400)(10,20283316)(11,0)(12,0)(13,70073220)(14,17685830)(15,45405782)(16,41199256)(17,42633484)(18,56683899)(19,25309217)(20,25030143)(21,22548371)(22,23330356)(23,37009794)};
		\addplot 
		coordinates {(1,3403743)(2,2235268)(3,1116111)(4,3096270)(5,3371448)(6,5156519)(7,5873153)(8,1336452)(9,6744349)(10,999601)(11,0)(12,0)(13,1123311)(14,4002828)(15,7250907)(16,5356971)(17,6495717)(18,5627376)(19,4812243)(20,790590)(21,681118)(22,1194899)(23,998105)};
		\legend{Added, Removed}
		\end{axis}
		\end{tikzpicture}
\caption{Evolution of total triples in Wikidata (left) and number of triples added to and removed from the $n$\textsuperscript{th} version for the $n+1$\textsuperscript{th} version (right) \label{fig:wikidata}}
\end{figure}

%\subsection{Queries Dataset}

\paragraph{Queries:} In order to achieve a set of SPARQL queries that are answerable over Wikidata and with which we could run experiments, we took the user-contributed example queries from the Wikidata Query Service\footnote{https://www.wikidata.org/wiki/Wikidata:SPARQL\_query\_service/queries/examples}, consisting of 388 SPARQL 1.1 \texttt{SELECT} queries of varying degrees of complexity, touching upon various domains of data, and with a mix of varying query operators.\footnote{Very recently, a large log of millions of SPARQL queries have been released for Wikidata~\cite{MalyshevKGGB18}; these queries were not available when conducting our experiments, but would be interesting to analyze as part of future work.}

Certain queries had to be removed from the original set of 388: some had syntax errors; others asked for information external to Wikidata (using \texttt{SERVICE}) or asked for qualifiers that are not present in the truthy version. We also eliminated a handful of queries that returned bindings to blank nodes to facilitate comparison between the results. Furthermore, we needed to transform certain queries for the purposes of our experiments. Some queries used features particular to the Wikidata Query Service, such as using a custom label \texttt{SERVICE} call to specify language preferences; we converted these queries to standard SPARQL using \texttt{COALESCE} to capture preferences. Furthermore, some queries featured non-deterministic elements -- such as \texttt{LIMIT}/\texttt{OFFSET} without ordering, \texttt{SAMPLE} or temporal functions -- such as \texttt{TODAY()} -- that may lead to changes in results not related to changes in the data; we remove queries with \texttt{SAMPLE} and temporal functions, and add an \texttt{ORDER BY} clause to all queries to ensure determinism with \texttt{LIMIT}/\texttt{OFFSET} and more generally to facilitate quick comparison of results. Figure~\ref{fig:transform} shows an example of a Wikidata query with the \texttt{LABEL} service rewritten and a deterministic ordering added. As a result of this process of filtering and transformation, we end up with a total of 221 deterministic queries.

\begin{figure}[!t]
\centering
{
	\begin{minipage}{0.8\textwidth}
	\begin{lstlisting}[style=sparqld]
#Cats
SELECT ?item ?itemLabel
WHERE {
 ?item wdt:P31 wd:Q146 .
 SERVICE wikibase:label { bd:serviceParam wikibase:language "en,es" }
}
	\end{lstlisting}
	\end{minipage}	
}

{
\begin{minipage}{0.8\textwidth}
	\begin{lstlisting}[style=sparqld]
#Cats
SELECT ?item ?itemLabel
WHERE {
  ?item wdt:P31 wd:Q146 .
  OPTIONAL { ?item rdfs:label ?en . FILTER(lang(?en)="en") }
  OPTIONAL { ?item rdfs:label ?es . FILTER(lang(?es)="es") }
  BIND(str(coalesce(?en, ?es, strafter(str(?item),
    "http://www.wikidata.org/entity/"))) AS ?itemLabel )
}
ORDER BY ?item ?itemLabel
	\end{lstlisting}
\end{minipage}	
}
\caption{Query transformation before (above) and after (below) \label{fig:transform}}
\end{figure}

In Table~\ref{tab:Pattern}, we provide statistics on the distribution of different types of triple patterns in the 221 queries; of key importance is that 90.35\% of the triple patterns have a constant in the predicate position, meaning that they are compatible with a model based on the dynamics of predicates. We also see that 83 triple patterns (7.93\%) feature a path expression, where Table~\ref{tab:Path} provides the distribution of these expressions (remarking that multiple patterns can be used in one expression); of note here is that recursion is commonly found (74.70\% use $e\texttt{*}$ while 9.64\% use $e\texttt{+}$), and that we find no negated property paths.

\begin{table}[t]
	\begin{minipage}[b]{0.47\linewidth}
		\centering
		\caption{Distribution of triple patterns (C: Constant, V: Variable, P: Path). \label{tab:Pattern}}
		\begin{tabular}{lrr}    \toprule
			\textbf{Pattern} & \textbf{\ntextnumero} & \textbf{\%} \\    \midrule
			V C V   & 751      & 71.73\%    \\
			V C C   & 189      & 18.05\%    \\
			V V V   & 9        & 0.86\%     \\
			C C V   & 6        & 0.57\%     \\
			V V C   & 5        & 0.48\%     \\
			C V V   & 3        & 0.29\%     \\
			C V C   & 1        & 0.10\%      \\
			C C C   & 0        & 0.00\%        \\    \midrule
			V P C   & 69       & 6.59\%     \\
			V P V   & 13       & 1.24\%     \\
			C P V   & 1        & 0.10\%      \\
			C P C   & 0        & 0.00\%        \\    \bottomrule
		\end{tabular}
	\end{minipage}
	\hfill
	\begin{minipage}[b]{0.47\linewidth}
		\centering
		\caption{Distribution of paths\label{tab:Path}}
		\begin{tabular}{lrr}    \toprule
			\textbf{Pattern}      & \textbf{\ntextnumero} & \textbf{\%} \\    \midrule
			$e\texttt{*}$   & 62       & 74.70\%     \\
			$e_1 \texttt{/} e_2$  & 56       & 67.47\%    \\
			$e\texttt{+} $   & 8        & 9.64\%     \\
			$e_1 \texttt{|} e_2$  & 6        & 7.23\%     \\
			$e\texttt{?}$    & 6        & 7.23\%     \\
			$\texttt{!(}e\texttt{)}$ & 0        & 0.00\%        \\ \midrule
			$e$ & 83 & 100.00\% \\
			\bottomrule
		%	$P \rightarrow C $    & 83       & 100.00\%      \\    \bottomrule
		\end{tabular}
	\end{minipage}
\end{table}

Finally, we look at how the results for these queries change over the 23 weeks. Figure~\ref{graph:changeV} shows how many queries had some results change between two consecutive versions, where we see that approximately half of the queries see changes in their results each week. But are these always the same queries that change every week? Figure~\ref{graph:changeV} shows the number of weeks in which some result changed for each query; the queries are ordered by the number of changes in their results, where we can see that the results of 44 queries (19.90\%) change each time, 18 queries (8\%) never change, and more generally, we note a quite uniform distribution of queries in between. More generally speaking, we conclude that our query set has a good balance of queries whose results never change, queries whose results always change, and queries whose results sometimes change.

\begin{figure}[t]
	\begin{minipage}[b]{0.47\linewidth}
		\centering
		\begin{tikzpicture}[thick,font=\scriptsize]
		\begin{axis}[
		xlabel={Version},
		ylabel={Queries changed},
		ylabel near ticks,
		width=6.5cm, 
		height=5cm,
		ymin=0, ymax=221,
		xtick={1,2,...,23},
		xticklabels={},
		extra x ticks={1,23},
		legend pos=north west,
		ymajorgrids=true,
		grid style=dashed,
		]

@@

		\end{axis}
		\end{tikzpicture}
		\caption{Number of queries with some results changed for each version}
		\label{graph:changeV}
	\end{minipage}
	\hspace{0.7cm}
	\begin{minipage}[b]{0.47\linewidth}
		\centering
		\begin{tikzpicture}[thick,font=\scriptsize]
		\begin{axis}[
		xlabel={Query},
		ylabel={Versions changed},
		ylabel near ticks,
		width=6.5cm, 
		height=5cm,
		ymin=0,
		legend pos=north west,
		ymajorgrids=true,
		grid style=dashed,
		]

@@

		\end{axis}
		\end{tikzpicture}
		\caption{Number of versions in which some results changed for each query}
		\label{graph:changeH}
	\end{minipage}
\end{figure}
